---
title: "Opioids Data Question - Coral Snakes"
author: "Amanuel Sidamo, Branden Dahlem, David Haines"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

rm(list=ls())
# memory.size(max=F)
setwd('.')
library(tidyverse)
library(ggplot2)
library(tidyr)
library(magrittr)
library(dplyr)
library(broom)
library(rmarkdown)
library(knitr)
library(RColorBrewer)
overdoses <- read_csv('./Data/overdoses.csv')
prescribers <- read_csv('./Data/prescriber-info.csv')
opioids <- read_csv('./Data/opioids.csv')
prescribers_2016 <- read_csv('./Data/prescriber-info_CY16.csv')
deaths_2016 <- read_csv('./Data/deaths-by-state-US-2016.csv')

colnames(overdoses)[colnames(overdoses) == 'State'] <- 'Full_State_Name'
colnames(overdoses)[colnames(overdoses) == 'Abbrev'] <- 'State'
colnames(deaths_2016)[colnames(deaths_2016) == 'number'] <- 'Deaths_2016'
colnames(deaths_2016)[colnames(deaths_2016) == 'State'] <- 'State_2016'

rx_per_2016 <- gather(prescribers_2016, rx, countrx, ACETAMINOPHEN.CODEINE:ZOLPIDEM.TARTRATE)
rx_per_prescriber <- gather(prescribers, rx, countrx, ABILIFY:ZOLPIDEM.TARTRATE)
prescriber <- rx_per_prescriber %>% 
              filter(Opioid.Prescriber == 1)
non_prescriber <- rx_per_prescriber %>% 
              filter(Opioid.Prescriber == 0)
prescriber_2016 <- rx_per_2016 %>% 
                    filter(Opioid.Prescriber == 1)
non_prescriber_2016 <- rx_per_2016 %>% 
                        filter(Opioid.Prescriber == 0)

drops <- c("PR", "ZZ", "AA", "AE", "GU", "VI", "DC", "AP", NA)
drop_states <- prescriber$State %in% drops
drop_deaths_2016 <- deaths_2016$State_2016 %in% drops
drop_states_2016 <- prescriber_2016$State %in% drops
non_drop_states <- non_prescriber$State %in% drops
non_drop_states_2016 <- non_prescriber_2016$State %in% drops
zero_value <- c(0)
drop_null_rx <- prescriber$countrx %in% zero_value
drop_null_rx_2016 <- prescriber_2016$countrx %in% zero_value
prescriber <- prescriber[drop_states == FALSE, ]
deaths_2016 <- deaths_2016[drop_deaths_2016 == FALSE, ]
prescriber_2016 <- prescriber_2016[drop_states_2016 == FALSE, ]
non_prescriber <- non_prescriber[non_drop_states == FALSE, ]
non_prescriber_2016 <- non_prescriber_2016[non_drop_states_2016 == FALSE, ]
prescriber_no_null <- prescriber[drop_null_rx == FALSE, ]
prescriber_no_null_2016 <- prescriber_2016[drop_null_rx_2016 == FALSE, ]
prescriber_by_rx <- prescriber %>% 
  group_by(rx)
overdoses = mutate(overdoses, Ratio = round(Deaths/Population * 100000, digits = 2))
deaths_2016 %<>% 
  mutate(Population = round(deaths_2016$Deaths_2016/(deaths_2016$rate/100000)))

colnames(deaths_2016)[colnames(deaths_2016) == 'Population'] <- 'Population_2016'

matches <- c("ACETAMINOPHEN.CODEINE", "FENTANYL", "HYDROCODONE.ACETAMINOPHEN", "HYDROMORPHONE.HCL","METHADONE.HCL","MORPHINE.SULFATE", "MORPHINE.SULFATE.ER", "OXYCODONE.HCL", "OXYCODONE.ACETAMINOPHEN","OXYCONTIN", "TRAMADOL.HCL")

op_match <- prescriber$rx %in% matches
opioids_prescribed <- rx_per_prescriber$rx %in% matches

op_match_2016 <- prescriber_2016$rx %in% matches

non_op_match <- non_prescriber$rx %in% matches

non_op_match_2016 <- non_prescriber_2016$rx %in% matches


list1 <- 1:25
list4 <- rep(('2016'), length(list1))
list2 <- rep(("2014"),length(list1))
list6 <- 1:549
list7 <- rep(('2014'), length(list6))
list8 <- 1:547
list9 <- rep(('2016'), length(list8))
rx_opioids <- rx_per_prescriber[opioids_prescribed == TRUE,]
```
## Introduction

While opioid pain relief drugs have been available for nearly 100 years, an "epidemic" of overdose-related deaths has been noted among public health officials and organizations in the last 20 years.

In this presentation, we analyzed publicly available datasets to identify trends in opioid prescription behavior from 2014.  The assumption of a Pareto effect was tested and multiple models were tested for predictive validity.  In addition, a dataset from 2016 was used to assess the predictive validity of the model and analyze trends in prescription practices based on observed results.

### Initial findings

An exploratory analysis resulted in multiple findings.  Some states had considerably greater rates of opioid prescription.  These rates were found to be consistent when controlling for state population.

```{r overall&stateplots, echo=FALSE}

drop_states <- prescriber$State %in% drops

drop_deaths_2016 <- deaths_2016$State_2016 %in% drops

drop_states_2016 <- prescriber_2016$State %in% drops

non_drop_states <- non_prescriber$State %in% drops

non_drop_states_2016 <- non_prescriber_2016$State %in% drops

prescriber <- prescriber[drop_states == FALSE, ]

deaths_2016 <- deaths_2016[drop_deaths_2016 == FALSE, ]

prescriber_2016 <- prescriber_2016[drop_states_2016 == FALSE, ]

non_prescriber <- non_prescriber[non_drop_states == FALSE, ]

non_prescriber_2016 <- non_prescriber_2016[non_drop_states_2016 == FALSE, ]

prescriber_by_rx <- prescriber %>% 
                    group_by(rx)

#op_match <- prescriber[op_match == TRUE, ]

#op_match_2016 <- prescriber[op_match == TRUE, ]

#op_mismatch <- prescriber[op_match != TRUE, ]
```
```{r debug11, echo=FALSE}
#op_mismatch_2016 <- prescriber_2016[op_match != TRUE, ]

#non_op_match <- non_prescriber[non_op_match == TRUE, ]

#non_op_match_2016 <- non_prescriber_2016[non_op_match_2016 == TRUE, ]

#op_match_by_state <- aggregate(data.frame(count = op_match$State), list(value = op_match$State), length)

#op_match_by_state_2016 <- aggregate(data.frame(count = op_match_2016$State), list(value = op_match_2016$State), length)

#op_match_by_rx <- aggregate(data.frame(count = op_match$rx), list(value = op_match$rx), length)

#op_match_by_state <- as.data.frame(op_match_by_state)

#op_match_by_state_2016 <- as.data.frame(op_match_by_state_2016)

#op_match_by_state$count <- as.numeric(op_match_by_state$count)

#op_match_by_state_2016$count <- as.numeric(op_match_by_state_2016$count)

#count_by_state <- op_match_by_state %>% 
#  group_by(value) %>% 
#  summarise(total = sum(count))

#count_by_state_2016 <- op_match_by_state_2016 %>% 
#  group_by(value) %>% 
#  summarise(total = sum(count))
```

```{r debug9, echo=FALSE}

#The dataset listed below is cleaned to include ONLY valid US states, non-null values for 'countrx', and prescription drugs listed as opioids
overdose_df <- overdoses
overdose_df_SORTED<-overdose_df[order(overdose_df$Deaths,decreasing=TRUE),]

sorted_deaths_plot <-  ggplot(data=overdose_df_SORTED, aes(x=State, y=Deaths,fill='red')) +
  geom_bar(stat="identity") +
  guides(fill=FALSE,color=FALSE) +
  coord_flip()

op_count_TX <- op_count_by_state %>% 
  filter(State == 'TX')

op_count_NH <- op_count_by_state %>% 
  filter(State == 'NH')

op_count_MT <- op_count_by_state %>% 
  filter(State == 'MT')

op_count_RI <- op_count_by_state %>% 
  filter(State == 'RI')

op_count_WV <- op_count_by_state %>% 
  filter(State == 'WV')

op_count_WV = mutate(op_count_WV, Ratio = total / sum(total))

op_count_NM <- op_count_by_state %>% 
  filter(State == 'NM')

op_count_KY <- op_count_by_state %>% 
  filter(State == 'KY')
```

Texas, Montana, New Hampshire, Rhode Island, West Virginia, New Mexico and Kentucky were among the top states for overall prescription counts.  Across these states, Hydrocodone/Acetaminophen (Lortab) was the most commonly prescribed opioid.

```{r stateplots, echo=FALSE}
ggplot(op_count_TX, aes(x = reorder(rx, total), y = total)) + geom_col() + labs(x = "Drug Names", y = "Total Prescriptions") + theme(axis.text.x = element_text(angle = 45, hjust = 1)) + ggtitle('Texas Prescription Count')

ggplot(op_count_NH, aes(x = reorder(rx, total), y = total)) + geom_col() + labs(x = "Drug Names", y = "Total Prescriptions") + theme(axis.text.x = element_text(angle = 45, hjust = 1)) + ggtitle('New Hampshire Prescription Count')

ggplot(op_count_MT, aes(x = reorder(rx, total), y = total)) + geom_col() + labs(x = "Drug Names", y = "Total Prescriptions") + theme(axis.text.x = element_text(angle = 45, hjust = 1)) + ggtitle('Montana Prescription Count')

ggplot(op_count_RI, aes(x = reorder(rx, total), y = total)) + geom_col() + labs(x = "Drug Names", y = "Total Prescriptions") + theme(axis.text.x = element_text(angle = 45, hjust = 1)) + ggtitle('Rhode Island Prescription Count')

ggplot(op_count_WV, aes(x = reorder(rx, total), y = total)) + geom_col() + labs(x = "Drug Names", y = "Total Prescriptions") + theme(axis.text.x = element_text(angle = 45, hjust = 1)) + ggtitle('West Virginia Prescription Count')

ggplot(op_count_NM, aes(x = reorder(rx, total), y = total)) + geom_col() + labs(x = "Drug Names", y = "Total Prescriptions") + theme(axis.text.x = element_text(angle = 45, hjust = 1)) + ggtitle('New Mexico Prescription Count')

ggplot(op_count_KY, aes(x = reorder(rx, total), y = total)) + geom_col() + labs(x = "Drug Names", y = "Total Prescriptions") + theme(axis.text.x = element_text(angle = 45, hjust = 1)) + ggtitle('Kentucky Prescription Count')

opioid_plot_1 <- ggplot() + 
  geom_bar(aes(y = Deaths, x = State, fill = Ratio), data = overdoses, stat="identity") + 
  labs(x = "State", y = "Deaths", fill = 'Deaths Per 100,000') + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  ggtitle('Overdoses by State')
 

opioid_plot_2 <- ggplot(data = overdoses,aes(x = State, y = Deaths)) + 
  geom_bar(aes(fill = factor(round(100000*Ratio))), stat = 'identity') + 
  scale_colour_manual(values = (brewer.pal(3,"OrRd")), aesthetics = 'fill') +
  labs(x = "State", y = "Deaths", fill = 'Deaths Per 100,000') + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  ggtitle('Overdoses by State')

colnames(deaths_2016)[colnames(deaths_2016) == 'Ratio'] <- 'Ratio_2016'

opioid_plot_2016 <- ggplot() + 
  geom_bar(aes(y = Deaths_2016, x = State_2016, fill = 100000*Ratio_2016), data = deaths_2016, stat="identity") + 
  labs(x = "State", y = "Deaths", fill = 'Deaths Per 100,000') + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  ggtitle('Overdoses by State 2016')

```

## Testing assumptions of the Pareto Effect:

The Pareto effect states that, in a given population with an outcome of interest, 20% of the population will be responsible for 80% of the effect on the given outcome.  In our scenario, we tested whether 20% of prescribers were responsible for 80% of the prescriptions issued in the US.

```{r, echo=FALSE}
rx_opioids<-rx_opioids[order(rx_opioids$countrx,decreasing=TRUE),]


tot_prescription<-sum(rx_opioids$countrx,na.rm = TRUE)
#print(paste0("Total Opioid Prescriptions: ", tot_prescription))
```


```{r, echo=FALSE}
rx_opioids$pctPrescribed<- (rx_opioids$countrx/tot_prescription)*100
rx_opioids_per_state<-data.frame(rx_opioids$State,rx_opioids$countrx,rx_opioids$pctPrescribed)
colnames(rx_opioids_per_state)<-c('state','countrx','pctPrescribed')
rx_opioids_per_state<-rx_opioids_per_state%>%
  group_by(state)%>%
  summarise(countrx=sum(countrx),pctPrescribed=sum(pctPrescribed))
rx_opioids_per_state<-rx_opioids_per_state[order(rx_opioids_per_state$pctPrescribed,decreasing = FALSE),]
```

Observing the cumulative sum of prescriptions shows

```{r, echo=FALSE}

rx_opioids_per_state$cumulativePct<-cumsum(rx_opioids_per_state$pctPrescribed)

```

```{r debug1, echo=FALSE}

#===================AMANUEL====================#

opioids_df <- opioids
overdose_df <- overdoses
overdose_df_SORTED<-overdose_df[order(overdose_df$Deaths,decreasing=TRUE),]


opioids_prescribed <- rx_per_prescriber$rx %in% matches
rx_opioids <- rx_per_prescriber[opioids_prescribed == TRUE,]
drop_states <- rx_opioids$State %in% drops
rx_opioids <- rx_opioids[drop_states == FALSE, ]
rx_opioids<- filter(rx_opioids,Opioid.Prescriber==1,!(is.na(countrx)))


top_opioids<-rx_opioids%>%
mutate(HYDROCODONE.ACETAMINOPHEN = if_else(`rx` == 'HYDROCODONE.ACETAMINOPHEN', 1, 0),
       TRAMADOL.HCL = if_else(`rx` == 'TRAMADOL.HCL', 1, 0),
       MORPHINE.SULFATE.ER = if_else(`rx` == 'MORPHINE.SULFATE.ER', 1, 0),
       OXYCODONE.HCL = if_else(`rx` == 'OXYCODONE.HCL', 1, 0),
       OXYCODONE.ACETAMINOPHEN = if_else(`rx` == 'OXYCODONE.ACETAMINOPHEN', 1, 0)) 

top5 <- top_opioids%>%
  mutate(HYDROCODONE.ACETAMINOPHEN=countrx*HYDROCODONE.ACETAMINOPHEN,
         TRAMADOL.HCL=countrx*TRAMADOL.HCL,
         MORPHINE.SULFATE.ER=countrx*MORPHINE.SULFATE.ER,
         OXYCODONE.HCL=countrx*OXYCODONE.HCL,
         OXYCODONE.ACETAMINOPHEN=countrx*OXYCODONE.ACETAMINOPHEN)



# bar plot number of overdoses related deaths by states
  ggplot(data=overdose_df_SORTED, aes(x=State, y=Deaths,fill='red')) +
  geom_bar(stat="identity") +
  guides(fill=FALSE,color=FALSE) +
  coord_flip()
# to plot the bar plot horizontally  coord_flip to plot verticall bars just run deaths_by_state


colnames(deaths_2016)[colnames(deaths_2016) == 'Population'] <- 'Population_2016'

rx_per_2016 <- gather(prescribers_2016, rx, countrx, ACETAMINOPHEN.CODEINE:ZOLPIDEM.TARTRATE)
rx_per_prescriber <- gather(prescribers, rx, countrx, ABILIFY:ZOLPIDEM.TARTRATE)

prescriber <- rx_per_prescriber %>% 
  filter(Opioid.Prescriber == 1)
non_prescriber <- rx_per_prescriber %>% 
  filter(Opioid.Prescriber == 0)

prescriber_2016 <- rx_per_2016 %>% 
  filter(Opioid.Prescriber == 1)
non_prescriber_2016 <- rx_per_2016 %>% 
  filter(Opioid.Prescriber == 0)
```

```{r, debug4, echo=FALSE}
prescribers <- read_csv('./Data/prescriber-info.csv')

rx_per_prescriber <- gather(prescribers, rx, countrx, ABILIFY:ZOLPIDEM.TARTRATE)

# create logical variable determines if RX by the NPIs are Opioids or not
opioids_prescribed <- rx_per_prescriber$rx %in% matches

# filter only NPI records where the prescribed drugs were opioids
rx_opioids <- rx_per_prescriber[opioids_prescribed == TRUE,]

# drop non_state USA territiries listed in the data

drops <- c("PR", "ZZ", "AA", "AE", "GU", "VI", "DC", "AP", NA)
drop_states <- rx_opioids$State %in% drops

# filter only valid states
rx_opioids <- rx_opioids[drop_states == FALSE, ]

# keep only qualified prescribers and countrx column not missing

rx_opioids<- filter(rx_opioids,Opioid.Prescriber==1,!(is.na(countrx)))

top_opioids<-rx_opioids%>%
mutate(HYDROCODONE.ACETAMINOPHEN = if_else(`rx` == 'HYDROCODONE.ACETAMINOPHEN', 1, 0),
       TRAMADOL.HCL = if_else(`rx` == 'TRAMADOL.HCL', 1, 0),
       MORPHINE.SULFATE.ER = if_else(`rx` == 'MORPHINE.SULFATE.ER', 1, 0),
       OXYCODONE.HCL = if_else(`rx` == 'OXYCODONE.HCL', 1, 0),
       OXYCODONE.ACETAMINOPHEN = if_else(`rx` == 'OXYCODONE.ACETAMINOPHEN', 1, 0)) 

top5 <- top_opioids%>%
  mutate(HYDROCODONE.ACETAMINOPHEN=countrx*HYDROCODONE.ACETAMINOPHEN,
         TRAMADOL.HCL=countrx*TRAMADOL.HCL,
         MORPHINE.SULFATE.ER=countrx*MORPHINE.SULFATE.ER,
         OXYCODONE.HCL=countrx*OXYCODONE.HCL,
         OXYCODONE.ACETAMINOPHEN=countrx*OXYCODONE.ACETAMINOPHEN )
```

```{r debug5, echo=FALSE}
# keep only State and top 5 Opioids columns

top5<-data.frame(top5$State,top5$HYDROCODONE.ACETAMINOPHEN,top5$TRAMADOL.HCL,top5$MORPHINE.SULFATE.ER,
 
                        top5$OXYCODONE.HCL,top5$OXYCODONE.ACETAMINOPHEN)

colnames(top5)<-c('State','HYDROCODONE.ACETAMINOPHEN','TRAMADOL.HCL','MORPHINE.SULFATE.ER',
                  
                  'OXYCODONE.HCL','OXYCODONE.ACETAMINOPHEN')

top5 %>%
group_by(State)%>%
summarize(HYDROCODONE.ACETAMINOPHEN=sum(HYDROCODONE.ACETAMINOPHEN),
          TRAMADOL.HCL=sum(TRAMADOL.HCL),
          MORPHINE.SULFATE.ER=sum(MORPHINE.SULFATE.ER),
          OXYCODONE.HCL=sum(OXYCODONE.HCL),
          OXYCODONE.ACETAMINOPHEN=sum(OXYCODONE.ACETAMINOPHEN))

  # question # 2 test pareto effect (20% of top prescribers are prescribing 80% of prescriptions)

# let's sort rx_opiodis dataset by countrx (count of presscriptions)
rx_opioids<-rx_opioids[order(rx_opioids$countrx,decreasing=TRUE),]
# let's find total opiodis prescription in by using 
tot_prescription<-sum(rx_opioids$countrx,na.rm = TRUE)

# calculate opioids precentage per drug per NPI 
rx_opioids$pctPrescribed<- (rx_opioids$countrx/tot_prescription)*100


# take state, countrx and pctPrescribed
rx_opioids_per_state<-data.frame(rx_opioids$State,rx_opioids$countrx,rx_opioids$pctPrescribed)
colnames(rx_opioids_per_state)<-c('state','countrx','pctPrescribed')
rx_opioids_per_state<-rx_opioids_per_state%>%
  group_by(state)%>%
  summarise(countrx=sum(countrx),pctPrescribed=sum(pctPrescribed))
rx_opioids_per_state<-rx_opioids_per_state[order(rx_opioids_per_state$pctPrescribed,decreasing = FALSE),]

rx_opioids_per_state$cumulativePct<-cumsum(rx_opioids_per_state$pctPrescribed)

ggplot(rx_opioids_per_state, aes(x=reorder(rx_opioids_per_state$state,rx_opioids_per_state$pctPrescribed))) +
  geom_bar(aes(y=rx_opioids_per_state$pctPrescribed), fill='blue', stat="identity") +
  geom_point(aes(y=rx_opioids_per_state$cumulativePct), color = rgb(0, 1, 0), pch=16, size=1) +
  geom_path(aes(y=rx_opioids_per_state$cumulativePct, group=1), colour="slateblue1", lty=3, size=0.9) + 
  theme(axis.text.x = element_text(angle=90, vjust=0.6)) + 
  labs(title = "Pareto Effect", x = 'states', y = 'pctPrescribed')
 theme(plot.title = element_text(hjust = 0.5))
```

```{r debug6, echo=FALSE}
# take only NPI, countrx and pctPrescribed columns

opioids_per_NPI<-data.frame(rx_opioids$NPI,rx_opioids$countrx,rx_opioids$pctPrescribed)

#  opioids per rx
opioids_per_rx<-data.frame(rx_opioids$rx,rx_opioids$countrx,rx_opioids$pctPrescribed)
# rename the columns as "NPI","countrx","pctPrescribed"
colnames(opioids_per_NPI)<-c('NPI','countrx',"pctPrescribed")
# rename columns
colnames(opioids_per_rx)<-c('rx','countrx',"pctPrescribed")

# uncomment the next line to compute for actual prescribers only.
#rx_opioids_per_NPI<-  filter(rx_opioids_per_NPI,countrx!=0)
# group by NPI 
opioids_per_NPI<-opioids_per_NPI%>% 
group_by(NPI)%>%
summarise(countrx=sum(countrx),pctPrescribed=sum(pctPrescribed))
# sort by descending order by percentage of opioids prescribed per NPI
#prescription_per_NPI<-prescription_per_NPI[order(prescription_per_NPI$pctPrescribed,decreasing=TRUE),]
# create cummulative sum of percentages prescribed by prescribers

opioids_per_NPI<-opioids_per_NPI[order(opioids_per_NPI$pctPrescribed,decreasing = TRUE),]
opioids_per_NPI$cumulativePct<-cumsum(opioids_per_NPI$pctPrescribed)


# group by rx 
opioids_per_rx<-opioids_per_rx%>% 
  group_by(rx)%>%
  summarise(countrx=sum(countrx),pctPrescribed=sum(pctPrescribed))


# this should be 100%
sum(opioids_per_NPI$pctPrescribed)


#group by RX

# sort by decreasing order

opioids_per_rx<-opioids_per_rx[order(opioids_per_rx$pctPrescribed,decreasing = TRUE),]
# print percentage share of countrx
print(opioids_per_rx)

opioids_per_rx$cumulativePct<-cumsum(opioids_per_rx$pctPrescribed)

ggplot(opioids_per_rx, aes(x=reorder(opioids_per_rx$rx,-opioids_per_rx$pctPrescribed))) +
  geom_bar(aes(y=opioids_per_rx$pctPrescribed), fill='blue', stat="identity") +
  geom_point(aes(y=opioids_per_rx$cumulativePct), color = rgb(0, 1, 0), pch=16, size=1) +
  geom_path(aes(y=opioids_per_rx$cumulativePct, group=1), colour="slateblue1", lty=3, size=0.9) + 
  theme(axis.text.x = element_text(angle=90, vjust=0.6)) + 
  labs(title = "Pareto Effect", x = 'opioidsPrescribed', y = 'pctPrescribed')
  theme(plot.title = element_text(hjust = 0.5))


# select 80% of prescriptions
prescription_80_pct<-opioids_per_NPI[opioids_per_NPI$cumulativePct<=80,]
# calculate count of top 80 percent prescribers
count_top_80_pct<-nrow(prescription_80_pct)

# how many prescribers are prescribing 80 percent of the prescription?


# calculate count of bottom 20 percent prescribers
bottom_20_pct<-opioids_per_NPI[opioids_per_NPI$cumulativePct>80,]
count_bottom_20_pct<-nrow(bottom_20_pct)
# How many of the prescribers share only 20 percent of total prescription

# Do 20 percent of prescribers prescribe 80 percent of the total prescription?

pareto_effect<-count_top_80_pct*100/(count_bottom_20_pct+count_top_80_pct)

# look opioids_per_NPI

# divide the data into five equal shares of 20% of cumulative percentage prescribed.
#  group based on cumulative percentage of count of prescriptions by NPI and get count for ranges
first_20_pct<-nrow(opioids_per_NPI[opioids_per_NPI$cumulativePct<20,])
second_20_pct<-nrow(opioids_per_NPI[opioids_per_NPI$cumulativePct>=20 & opioids_per_NPI$cumulativePct<40,])
third_20_pct<-nrow(opioids_per_NPI[opioids_per_NPI$cumulativePct>=40 & opioids_per_NPI$cumulativePct<60,])
fourth_20_pct<-nrow(opioids_per_NPI[opioids_per_NPI$cumulativePct>=60 & opioids_per_NPI$cumulativePct<80,])
fifth_20_pct<-nrow(opioids_per_NPI[opioids_per_NPI$cumulativePct>=80,])

# rename columns as range

pareto_df_1<-data.frame(first_20_pct,second_20_pct,third_20_pct,fourth_20_pct,fifth_20_pct)

# renaming the columns here help to plot the ranges in order from first 20% to fifth 20% of prescribers.

colnames(pareto_df_1)<-c('share1','share2','share3','share4','share5')

# calculate share of prescriptions by ranges
pareto_df_1<-pareto_df_1%>%
  mutate(share1=round(fifth_20_pct/nrow(opioids_per_NPI),3),share2=round(fourth_20_pct/nrow(opioids_per_NPI),3),
         share3=round(third_20_pct/nrow(opioids_per_NPI),3),share4=round(second_20_pct/nrow(opioids_per_NPI),3),
         share5=round(first_20_pct/nrow(opioids_per_NPI),3))
# look the data

pareto_df_1<-pareto_df_1*100

# gather the data from wide to long
pareto_df_1<-gather(pareto_df_1)

# rename column names
colnames(pareto_df_1)<-c('cumulativePctShareRx','pctOfPrescribers')

# plot the pareto_df_1 dataset 
# 80 % of the prescribers share only 20% of the prescription

ggplot(pareto_df_1, aes(cumulativePctShareRx, pctOfPrescribers)) +   
  geom_bar(aes(fill = pctOfPrescribers), position = "dodge", stat="identity") +
  guides(fill=FALSE,color=FALSE) +
  ggtitle('Opioids Prescribers by cumulative \n Share of Prescriptions')+
  theme(plot.title = element_text(hjust = 0.5))


# total number of prescribers

# count of bottom 20 pct and top 80 pct prescribers

pareto_df<-data.frame(count_bottom_20_pct,count_top_80_pct)

# calculate percentage share of  bottom 80 pct and top 20 pct prescribers out of total prescription
pareto_df2<-pareto_df%>%
  transmute(bottom_80_pct=round(count_top_80_pct/(count_top_80_pct+count_bottom_20_pct),3),top_20_pct=round(count_bottom_20_pct/(count_top_80_pct+count_bottom_20_pct),3))

# have a look at the data
#rename columns

colnames(pareto_df2)<-c('bottom 80 pct','top 20 pct')

pareto<-gather(pareto_df2)
# rename columns
colnames(pareto)<-c('ranges','pctPrescribed')

  
# plot percentage of prescriptions shared by bottom 80 pct and top 20 pct prescribers

ggplot(pareto, aes(ranges, pctPrescribed)) +   
  geom_bar(aes(fill = pctPrescribed), position = "dodge", stat="identity") +
    guides(fill=FALSE,color=FALSE) +
     ggtitle('Pareto Effect for top 20 Percent vs Bottom 80 \n Percent Opioids Prescribers')+
      theme(plot.title = element_text(hjust = 0.5))


# Only 24.1% of the qualified prescribers have prescribed 75.59% of Opioids prescriptions
# when only actuall prescribers are considered 36% of prescribers prescribed 76% of the Opioids

# plot histogram 
ggplot(
 data = opioids_per_NPI,
 aes(x = cumulativePct,fill='orange2')
) +
  geom_histogram(binwidth =20)+
 guides(fill=FALSE,color=FALSE) +
  ggtitle('Distribution of Opioids Prescriptions')
  
  
# rename columns
#colnames(prescription_per_NPI)<-c('countRxPerNPI','cumulativePct')

# plot cummulative  distribution

plot(opioids_per_NPI$cumulativePct, xlab = 'totCountRx', ylab = 'cumulativePct', main = 'Cumulative Distribution\nOpioids Prescriptions')

```

```{r debug7, echo=FALSE}
# Question # 3  Build a model to predict the number of fatal opioid overdoses
# try to create a model for a specific state

# import population by race data, this data takes only non hispanic white.

race<-read_csv('./Data/PEP_2014_PEPSR6H_with_ann.csv')

# import pecentage of uninsured population by states

insurance<-read_csv('./Data/hic04_acs.csv')

# total insured row is not needed, we wilter it out.
insurance <-insurance%>%
  filter(coverage!='Total')
  

# import percapita health expenditure by states 

percapita_HE<-read_csv('./Data/raw_data.csv',skip =2)

# keep only by states data, we don't need USA total.
percapita_HE<-percapita_HE%>%
  filter(state!='United States')

# let's merge supplementary data with prescription data
# check headers of the datasets 

# clean race data, let's keep 2014 estimate 
```

```{r debug8, echo=FALSE}
race<-race%>%
  filter(Year.id=='est72014',Sex.id=='totsex',race$`GEO.display-label`!="United States")
race_df<-data.frame(race$`GEO.display-label`,race$Hisp.id,race$totpop,race$wa,race$ba,race$ia,race$aa,race$na,race$tom)

# keep only non_hispanic race
race_df<-race_df%>%
  filter(race.Hisp.id=='nhisp')


# rename columns
colnames(race_df)<-c('state','nonhisp','totpop','wa','ba','ia','aa','na','tom')
race_df<-race_df<-race_df%>%
  mutate(state=toupper(state),wa=round(wa/totpop*100,2),ba=round(ba/totpop*100,2),ia=round(ia/totpop*100,2),aa=round(aa/totpop*100,2),na=round(na/totpop*100,2),tom=round(tom/totpop*100,2))



# rename columns for percentage of insurance coveragd and unisured population by states

colnames(insurance)<-c('state','Coverage_type','uninsured2016','uninsuredPct2014')
# let's keep only uninsured from insurance dataset
uninsured<-insurance%>%
  filter(Coverage_type=='Uninsured')

# rename columns for per_capita health expenditure
colnames(percapita_HE)<-c('state','perCapitaHealthExp')

# states in opiods_prescription dataset are in abbreviation, we cannot match with other data sets
#So we need to create state name and state abbriviation data

state_df<-data.frame(state.name,state.abb)
colnames(state_df)<-c('state_name','State')
# state_df doesn't include district of columbia.
# create state name and abbreviation for DC.
DC_state<-data.frame(state_name="District of Columbia",State="DC")
# bind it with state_df
state_df<-rbind.data.frame(state_df,DC_state)


# to check the header of rx_opioids dataset

# join state names  data with prescription data
rx_opioids_df<-rx_opioids%>%
   left_join(state_df,by='State')

# rename column names

colnames(rx_opioids_df)[colnames(rx_opioids_df)=='State']<-'State_code'
colnames(rx_opioids_df)[colnames(rx_opioids_df)=='state_name']<-'state'

# to check if DC has full state name in the dataset
rx_opioids_df%>%
  filter(State_code=='DC')


# capitalize states to match with other data sets
rx_opioids_df<-rx_opioids_df%>%
  mutate(state=toupper(state))

# take selected columns
rx_opioids_df<-data.frame(rx_opioids_df$state,rx_opioids_df$NPI,rx_opioids_df$countrx,rx_opioids_df$Opioid.Prescriber)
# group rx_opioids_df by state
prescription_by_state<-rx_opioids_df%>% 
  group_by(rx_opioids_df.state)%>%
  summarize(countrx=sum(rx_opioids_df.countrx),total_prescribers=sum(rx_opioids_df.Opioid.Prescriber))
  
# rename columns

colnames(prescription_by_state)<-c('state','countrx','total_prescribers')


# merge overdose data with prescription data

# rename columns
colnames(overdose_df)<-c('state','population','deaths','abbrev', 'rate')

overdose_df<-overdose_df%>%
  mutate(state=toupper(state))

# merge overdose data with opioids rx
merged_df1<- overdose_df %>%
  left_join(prescription_by_state,by='state')



# join health percentage of uninsured population by states
merged_df2<-merged_df1 %>%
  left_join(uninsured,by='state')

# capitalize state names

percapita_HE<-percapita_HE%>%
  mutate(state=toupper(state))

# join per capita health expenditure 
merged_df3<-merged_df2 %>%
  left_join(percapita_HE,by='state')

# merge race_df to the merged data

merged_df4<-merged_df3%>%
  left_join(race_df,by='state')


# normalize deaths and count of prescription by 100 people
merged_df4<-merged_df4%>%
  mutate(deaths=deaths/100,countrx=countrx/100)

# import unemployment data

unemployment<-read_csv('./Data/emp_unemployment.csv')

# merge unemployment rate data 
merged_df5<-inner_join(merged_df4,unemployment,by='state')



# import age data
age_df<-read_csv('./Data/age_data.csv')

# merge age data
merged_df6<-left_join(merged_df5,age_df,by='state')


# import median household income
median_hsld_income<-read_csv('./Data/Median Household Income by State.csv')

# merge  median income data
merged_df7<-left_join(merged_df6,median_hsld_income,by='state')

# import education data

education_df<-read_csv('./Data/education_highscool_or_more.csv')

# Merge education data 
merged_df7<-left_join(merged_df7,education_df,by='state')

```

``` {r debug12, echo=FALSE}
# create data frame for only variables that will be used in the model
modelVariables<-data.frame(merged_df7$deaths,merged_df7$countrx, merged_df7$medianAge,merged_df7$unemployment,merged_df7$uninsuredPct2014,merged_df7$perCapitaHealthExp,merged_df7$Education,merged_df7$Income,merged_df7$wa,merged_df7$ba)

colnames(modelVariables)<-c('deaths','countrx','medianAge','unemp','uninsured','HealthExp','Educ','Income','wa','ba')

# calculate correlation among the model variables
# race variables have opposite sign in the regression compared with correlation coefficient.
round(cor(modelVariables),2)

# only countrx as dependent
lm_countrx<-lm(formula=deaths~countrx,data=merged_df7)


glance(lm_countrx)

coeff_lm_countrx<-tidy(lm_countrx)
write.csv(coeff_lm_countrx,'coeff_lm_countrx.csv',row.names = FALSE)
glance_lm_countrx<-round(glance(lm_countrx),3)
write.csv(glance_lm_countrx,'glance_lm_countrx.csv',row.names = FALSE)

augumented_lm_countrx<-augment(lm_countrx)

augumented_lm_countrx%>%
  ggplot(aes(x=countrx))+
  geom_point(aes(y=deaths))+
  geom_line(aes(y=.fitted,color='red')) +
  guides(fill=FALSE,color=FALSE) +
  ggtitle("relationship between overdose deaths\n  and countrx by states")
theme(plot.title = element_text(hjust = 0.5))


# linear model with all possible variables we have to estimate the overdose deaths
linearModel <-lm(formula = deaths~countrx + medianAge + unemployment +
                   uninsuredPct2014+perCapitaHealthExp+ +uninsuredPct2014+Income+Education +wa + ba, data =merged_df7 )

# check the coefficients and fitness of the model
tidy(linearModel)
glance(linearModel)
coeff_lm<-tidy(linearModel)
glance<-glance(linearModel)
augmented_df<-augment(linearModel)

write.csv(coeff_lm,'coeff_lm.csv',row.names = FALSE)
write.csv(glance,'glance_lm.csv',row.names = FALSE)
# plot the model
plot(linearModel)


# only countrx, percentage of unisured population and white

lm_countrx_unins_wa<-lm(formula = deaths~countrx+uninsuredPct2014+wa,data=merged_df7)
tidy(lm_countrx_unins_wa)
glance(lm_countrx_unins_wa)

# only countrx, percentage of unisured population and black

lm_countrx_unins_ba<-lm(formula = deaths~countrx+uninsuredPct2014+ba,data=merged_df7)
tidy(lm_countrx_unins_ba)
glance(lm_countrx_unins_ba)

# linear model by removing three weakly correlated variables

lm_corr_vars <-lm(formula = deaths~countrx + Education + unemployment +
                    uninsuredPct2014 + wa + ba, data =merged_df7 )
tidy(lm_corr_vars)
glance(lm_corr_vars)


# linear model to estimate the overdose deaths
augmented_df%>%
  ggplot(aes(x=medianAge))+
  geom_point(aes(y=deaths))+
  geom_line(aes(y=.fitted,color='red')) + 
  guides(fill=FALSE,color=FALSE) + 
  ggtitle("relationship between overdose deaths\n  and median_age by states") +
  theme(plot.title = element_text(hjust = 0.5))


augmented_df%>%
  ggplot(aes(x=perCapitaHealthExp))+
  geom_point(aes(y=deaths))+
  geom_line(aes(y=.fitted,color='red')) + 
  guides(fill=FALSE,color=FALSE) + 
  ggtitle("overdose deaths and health expenditure \n by States") +
  theme(plot.title = element_text(hjust = 0.5))

augmented_df%>%
  ggplot(aes(x=uninsuredPct2014))+
  geom_point(aes(y=deaths))+
  geom_line(aes(y=.fitted,color='red')) + 
  guides(fill=FALSE,color=FALSE) + 
  ggtitle("overdose deaths and percentage of uninsured") +
  theme(plot.title = element_text(hjust = 0.5))

augmented_df%>%
  ggplot(aes(x=unemployment))+
  geom_point(aes(y=deaths))+
  geom_line(aes(y=.fitted,color='red')) + 
  guides(fill=FALSE,color=FALSE) + 
  ggtitle("overdose deaths and unemployment rate") +
  theme(plot.title = element_text(hjust = 0.5))

augmented_df%>%
  ggplot(aes(x=Education))+
  geom_point(aes(y=deaths))+
  geom_line(aes(y=.fitted,color='red')) + 
  guides(fill=FALSE,color=FALSE) + ggtitle("overdose deaths and percentage of \n Highschool and above by states") +
  theme(plot.title = element_text(hjust = 0.5))

augmented_df%>%
  ggplot(aes(x=Income))+
  geom_point(aes(y=deaths))+ 
    geom_line(aes(y=.fitted,color='red')) + 
    guides(fill=FALSE,color=FALSE) + ggtitle("overdose deaths and median household income by states")

```

Models assessed:
- Count of rx only
- All variables
- Variables correlated with deaths
- Proportions of top 5 opioids prescribed within a state
- Count of rx and proportion of uninsured population per state - Significant!






